---
title: "Hands-on Exercise 8: Local Measures of Spatial Autocorrelation"
author: "Lorielle Malveda"
date: "September 23, 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  freeze: true
  warning: false
---

# 1. OVERVIEW

**Local Measures of Spatial Autocorrelation (LMSA)** analyze the relationships between each observation and its surroundings, offering detailed insights into the spatial structure of data rather than providing a global summary. These measures are not mere summary statistics; they are specific scores that reveal the spatial dynamics within the dataset. Often, these local measures are related to their global counterparts, sometimes being components that together form the global statistic. For example, Local Indicators of Spatial Association (LISA) are directly linked to broader global metrics and can be seen as their disaggregated form. Another local measure, the **Getis-Ord’s Gi-statistics**, offers complementary perspectives or similar insights for data with geographical references.

In this practical exercise, you will learn to compute various Local Measures of Spatial Autocorrelation using the `spdep` package.

By the end of this session, you will be able to:

-   Import geospatial data using the relevant functions from the `sf` package,
-   Load a CSV file using the `readr` package,
-   Perform relational joins using the appropriate functions from the `dplyr` package,
-   Calculate Local Indicator of Spatial Association (LISA) statistics to identify clusters and outliers using the `spdep` package,
-   Determine hot spots and cold spots using Getis-Ord’s Gi-statistics with the `spdep` package, and
-   Visualize your analysis results using the `tmap` package.

# 2. GETTING STARTED

## 2.1 Objective - The Analytical Question

In spatial policy, a key development goal for local governments and planners is to ensure an equal distribution of development across the province.

The objective of this study is to apply appropriate spatial statistical methods to determine whether development is distributed geographically. If it is not, the next step is to investigate whether there is evidence of spatial clustering. If clustering is present, we will then seek to identify the locations of these clusters.

In this case study, we focus on analyzing the spatial pattern of a specific development indicator—GDP per capita—in Hunan Province, People's Republic of China.

## 2.2 Study Area and Dataset

Two data sets will be used in this hands-on exercise, which are:

1.  **Hunan Province Administrative Boundary Layer (at County Level).** This is a geospatial data set in ESRI shapefile format.
2.  **Hunan_2012.csv.** This csv file contains selected Hunan’s local development indicators in 2012.

## 2.3 Setting the Analytical Tools

This code chunk uses `p_load()` of the ***pacman*** package (stands for Package Manager) to check if the following packages are installed:

**sf:** used for importing and handling geospatial data in R.

**tidyverse:** mainly used for wrangling attribute data in R.

***spdep:*** used to compute spatial weights, global and local spatial autocorrelation statistics.

**tmap:** used to prepare cartographic quality chropleth map.

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse)
```

If available, the packages will then be launched into R.

# 3. GETTING THE DATA INTO R ENVIRONMENT

In this section, you will learn how to bring a geospatial data and its associated attribute table into R environment. The geospatial data is in ESRI shapefile format and the attribute table is in csv fomat.

## 3.1 Importing *shapefile* into the R Environment

Let us use the ***sf*** package to import the Hunan shapefile into R.

The output is a simple features ***sf*** object.

```{r}
hunan <- st_read(dsn = "data/geospatial",                   layer = "Hunan")
```

## 3.2 Importing *csv* into the R Environment

Next, we will import `Hunan_2012.csv` into R by using the `read_csv` function of the ***readr*** package.

The output is R dataframe class.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

## 3.3 Performing Relational Join

The following code chunk updates the attribute table of the `hunan` SpatialPolygonsDataFrame by merging it with the attribute fields of the `hunan2012` dataframe. This is accomplished using the `left_join()` function from the `dplyr` package:

```{r}
hunan <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```

## 3.4 Visualizing the Regional Development Indicator

Next, we'll prepare a basemap and a choropleth map to display the distribution of GDP per capita for 2012, using the `qtm()` function from the `tmap` package.

```{r}
equal <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal quantile classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```

# 4. LOCAL INDICATORS OF SPATIAL ASSOCIATION (LISA)

Local Indicators of Spatial Association or LISA are statistics that evaluate the existence of clusters and/or outliers in the spatial arrangement of a given variable. For instance if we are studying distribution of GDP per capita of Hunan Provice, People Republic of China, local clusters in GDP per capita mean that there are counties that have higher or lower rates than is to be expected by chance alone; that is, the values occurring are above or below those of a random distribution in space.

In this section, you will learn how to apply appropriate Local Indicators for Spatial Association (LISA), especially local Moran’I to detect cluster and/or outlier from GDP per capita 2012 of Hunan Province, PRC.

## 4.1 Computing Contiguity Spatial Weights

Before we can compute the local spatial autocorrelation statistics, we need to construct a spatial weights of the study area. The spatial weights is used to define the neighbourhood relationships between the geographical units (i.e. county) in the study area.

In the code chunk below, [`poly2nb()`](https://r-spatial.github.io/spdep/reference/poly2nb.html) of **spdep** package is used to compute contiguity weight matrices for the study area. This function builds a neighbours list based on regions with contiguous boundaries. If you look at the documentation you will see that you can pass a “queen” argument that takes TRUE or FALSE as options. If you do not specify this argument the default is set to TRUE, that is, if you don’t specify queen = FALSE this function will return a list of first order neighbours using the Queen criteria.

More specifically, the code chunk below is used to compute Queen contiguity weight matrix.

```{r}
wm_q <- poly2nb(hunan, 
                queen=TRUE)
summary(wm_q)
```

The summary report above shows that there are 88 area units in Hunan. The most connected area unit has 11 neighbours. There are two area units with only one neighbours.

## 4.2 Row-standardized Weights Matrix

Next, we need to assign weights to each neighboring polygon. In our case, each neighboring polygon will be assigned equal weight (style=“W”). This is accomplished by assigning the fraction 1/(#ofneighbors) to each neighboring county then summing the weighted income values. While this is the most intuitive way to summaries the neighbors’ values it has one drawback in that polygons along the edges of the study area will base their lagged values on fewer polygons thus potentially over- or under-estimating the true nature of the spatial autocorrelation in the data. For this example, we’ll stick with the style=“W” option for simplicity’s sake but note that other more robust options are available, notably style=“B”.

```{r}
rswm_q <- nb2listw(wm_q, 
                   style="W", 
                   zero.policy = TRUE)
rswm_q
```

**What can we learn from the code chunk above?**

-   The input of [`nb2listw()`](https://r-spatial.github.io/spdep/reference/nb2listw.html) must be an object of class **nb**. The syntax of the function has two major arguments, namely style and zero.poly.

-   *style* can take values “W”, “B”, “C”, “U”, “minmax” and “S”. B is the basic binary coding, W is row standardised (sums over all links to n), C is globally standardised (sums over all links to n), U is equal to C divided by the number of neighbours (sums over all links to unity), while S is the variance-stabilizing coding scheme proposed by Tiefelsdorf et al. 1999, p. 167-168 (sums over all links to n).

-   If *zero policy* is set to TRUE, weights vectors of zero length are inserted for regions without neighbour in the neighbours list. These will in turn generate lag values of zero, equivalent to the sum of products of the zero row t(rep(0, length=length(neighbours))) %\*% x, for arbitrary numerical vector x of length length(neighbours). The spatially lagged value of x for the zero-neighbour region will then be zero, which may (or may not) be a sensible choice.

## 4.3 Computing Local Moran's I

To compute local Moran’s I, the [*localmoran()*](https://r-spatial.github.io/spdep/reference/localmoran.html) function of **spdep** will be used. It computes *Ii* values, given a set of *zi* values and a listw object providing neighbour weighting information for the polygon associated with the zi values.

The code chunks below are used to compute local Moran’s I of *GDPPC2012* at the county level.

```{r}
fips <- order(hunan$County)
localMI <- localmoran(hunan$GDPPC, rswm_q)
head(localMI)
```

*localmoran()* function returns a matrix of values whose columns are:

-   Ii: the local Moran’s I statistics

-   E.Ii: the expectation of local moran statistic under the randomisation hypothesis

-   Var.Ii: the variance of local moran statistic under the randomisation hypothesis

-   Z.Ii:the standard deviate of local moran statistic

-   Pr(): the p-value of local moran statistic

The code chunk below list the content of the local Moran matrix derived by using [*printCoefmat()*](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/printCoefmat).

```{r}
printCoefmat(data.frame(
  localMI[fips,], 
  row.names=hunan$County[fips]),
  check.names=FALSE)
```

### MAPPING THE LOCAL MORAN'S I

Before mapping the local Moran’s I map, it is wise to append the local Moran’s I dataframe (i.e. localMI) onto hunan SpatialPolygonDataFrame. The code chunks below can be used to perform the task. The out SpatialPolygonDataFrame is called *hunan.localMI*.

```{r}
hunan.localMI <- cbind(hunan,localMI) %>%
  rename(Pr.Ii = Pr.z....E.Ii..)
```

### MAPPING LOCAL MORAN'S I VALUES

Using choropleth mapping functions of **tmap** package, we can plot the local Moran’s I values by using the code chinks below.

```{r}
tm_shape(hunan.localMI) +
  tm_fill(col = "Ii", 
          style = "pretty",
          palette = "RdBu",
          title = "local moran statistics") +
  tm_borders(alpha = 0.5)
```

### MAPPING LOCAL MORAN's I P-VALUES

The choropleth shows there is evidence for both positive and negative Ii values. However, it is useful to consider the p-values for each of these values, as consider above.

The code chunks below produce a choropleth map of Moran’s I p-values by using functions of **tmap** package.

```{r}
tm_shape(hunan.localMI) +
  tm_fill(col = "Pr.Ii", 
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette="-Blues", 
          title = "local Moran's I p-values") +
  tm_borders(alpha = 0.5)
```

### MAPPING BOTH LOCAL MORAN'S I AND P-VALUES

For effective interpretation, it is better to plot both the local Moran’s I values map and its corresponding p-values map next to each other.

The code chunk below will be used to create such visualisation.

```{r}
localMI.map <- tm_shape(hunan.localMI) +
  tm_fill(col = "Ii", 
          style = "pretty", 
          title = "local moran statistics") +
  tm_borders(alpha = 0.5)

pvalue.map <- tm_shape(hunan.localMI) +
  tm_fill(col = "Pr.Ii", 
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette="-Blues", 
          title = "local Moran's I p-values") +
  tm_borders(alpha = 0.5)

tmap_arrange(localMI.map, pvalue.map, asp=1, ncol=2)
```

# 5. CREATING A *LISA* CLUSTER MAP

The LISA Cluster Map shows the significant locations color coded by type of spatial autocorrelation. The first step before we can generate the LISA cluster map is to plot the Moran scatterplot.

## 5.1 Plotting Moran Scatterplot

The Moran scatterplot is an illustration of the relationship between the values of the chosen attribute at each location and the average value of the same attribute at neighboring locations.

The code chunk below plots the Moran scatterplot of GDPPC 2012 by using [*moran.plot()*](https://r-spatial.github.io/spdep/reference/moran.plot.html) of **spdep**.

```{r}
nci <- moran.plot(hunan$GDPPC, rswm_q,
                  labels=as.character(hunan$County), 
                  xlab="GDPPC 2012", 
                  ylab="Spatially Lag GDPPC 2012")
```

Notice that the plot is split in 4 quadrants. The top right corner belongs to areas that have high GDPPC and are surrounded by other areas that have the average level of GDPPC. This are the high-high locations in the lesson slide.

## 5.2 Plotting Moran Scatterplot with Standardized Variable

First we will use [*scale()*](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/scale) to centers and scales the variable. Here centering is done by subtracting the mean (omitting NAs) the corresponding columns, and scaling is done by dividing the (centered) variable by their standard deviations.

```{r}
hunan$Z.GDPPC <- scale(hunan$GDPPC) %>% 
  as.vector 
```

The [*as.vector()*](https://www.rdocumentation.org/packages/pbdDMAT/versions/0.5-1/topics/as.vector) added to the end is to make sure that the data type we get out of this is a vector, that map neatly into out dataframe.

Now, we are ready to plot the Moran scatterplot again by using the code chunk below.

```{r}
nci2 <- moran.plot(hunan$Z.GDPPC, rswm_q,
                   labels=as.character(hunan$County),
                   xlab="z-GDPPC 2012", 
                   ylab="Spatially Lag z-GDPPC 2012")
```

## 5.3 Preparing LISA Map Classes

The code chunks below show the steps to prepare a LISA cluster map.

```{r}
quadrant <- vector(mode="numeric",length=nrow(localMI))
```

Next, derives the spatially lagged variable of interest (i.e. GDPPC) and centers the spatially lagged variable around its mean.

```{r}
hunan$lag_GDPPC <- lag.listw(rswm_q, hunan$GDPPC)
DV <- hunan$lag_GDPPC - mean(hunan$lag_GDPPC)     
```

This is follow by centering the local Moran’s around the mean.

```{r}
LM_I <- localMI[,1] - mean(localMI[,1])    
```

Next, we will set a statistical significance level for the local Moran.

```{r}
signif <- 0.05       
```

These four command lines define the low-low (1), low-high (2), high-low (3) and high-high (4) categories.

```{r}
quadrant[DV <0 & LM_I>0] <- 1
quadrant[DV >0 & LM_I<0] <- 2
quadrant[DV <0 & LM_I<0] <- 3  
quadrant[DV >0 & LM_I>0] <- 4      
```

Lastly, places non-significant Moran in the category 0.

```{r}
quadrant[localMI[,5]>signif] <- 0
```

In fact, we can combined all the steps into one single code chunk as shown below:

```{r}
quadrant <- vector(mode="numeric",length=nrow(localMI))
hunan$lag_GDPPC <- lag.listw(rswm_q, hunan$GDPPC)
DV <- hunan$lag_GDPPC - mean(hunan$lag_GDPPC)     
LM_I <- localMI[,1]   
signif <- 0.05       
quadrant[DV <0 & LM_I>0] <- 1
quadrant[DV >0 & LM_I<0] <- 2
quadrant[DV <0 & LM_I<0] <- 3  
quadrant[DV >0 & LM_I>0] <- 4    
quadrant[localMI[,5]>signif] <- 0
```

## 5.4 Plotting LISA map

Now, we can build the LISA map by using the code chunks below.

```{r}
hunan.localMI$quadrant <- quadrant
colors <- c("#ffffff", "#2c7bb6", "#abd9e9", "#fdae61", "#d7191c")
clusters <- c("insignificant", "low-low", "low-high", "high-low", "high-high")

tm_shape(hunan.localMI) +
  tm_fill(col = "quadrant", 
          style = "cat", 
          palette = colors[c(sort(unique(quadrant)))+1], 
          labels = clusters[c(sort(unique(quadrant)))+1],
          popup.vars = c("")) +
  tm_view(set.zoom.limits = c(11,17)) +
  tm_borders(alpha=0.5)
```

For effective interpretation, it is better to plot both the local Moran’s I values map and its corresponding p-values map next to each other.

The code chunk below will be used to create such visualisation.

```{r}
gdppc <- qtm(hunan, "GDPPC")

hunan.localMI$quadrant <- quadrant
colors <- c("#ffffff", "#2c7bb6", "#abd9e9", "#fdae61", "#d7191c")
clusters <- c("insignificant", "low-low", "low-high", "high-low", "high-high")

LISAmap <- tm_shape(hunan.localMI) +
  tm_fill(col = "quadrant", 
          style = "cat", 
          palette = colors[c(sort(unique(quadrant)))+1], 
          labels = clusters[c(sort(unique(quadrant)))+1],
          popup.vars = c("")) +
  tm_view(set.zoom.limits = c(11,17)) +
  tm_borders(alpha=0.5)

tmap_arrange(gdppc, LISAmap, 
             asp=1, ncol=2)
```

We can also include the local Moran’s I map and p-value map as shown below for easy comparison.

> Question: What statistical observations can you draw from the LISA map above?

# **6. Hot Spot and Cold Spot Area Analysis**

Beside detecting cluster and outliers, localised spatial statistics can be also used to detect hot spot and/or cold spot areas.

The term ‘hot spot’ has been used generically across disciplines to describe a region or value that is higher relative to its surroundings (Lepers et al 2005, Aben et al 2012, Isobe et al 2015).

## 6.1 Getis and Ord’s G-Statistics

An alternative spatial statistics to detect spatial anomalies is the Getis and Ord’s G-statistics (Getis and Ord, 1972; Ord and Getis, 1995). It looks at neighbours within a defined proximity to identify where either high or low values clutser spatially. Here, statistically significant hot-spots are recognised as areas of high values where other areas within a neighbourhood range also share high values too.

The analysis consists of three steps:

-   Deriving spatial weight matrix

-   Computing Gi statistics

-   Mapping Gi statistics

## 6.2 Deriving Distance-Based Weight Matrix

First, we need to define a new set of neighbours. Whist the spatial autocorrelation considered units which shared borders, for Getis-Ord we are defining neighbours based on distance.

There are two type of distance-based proximity matrix, they are:

-   fixed distance weight matrix; and

-   adaptive distance weight matrix.

### DERIVING THE CENTROID MATRIX

We will need points to associate with each polygon before we can make our connectivity graph. It will be a little more complicated than just running *st_centroid()* on the sf object: **us.bound**. We need the coordinates in a separate data frame for this to work. To do this we will use a mapping function. The mapping function applies a given function to each element of a vector and returns a vector of the same length. Our input vector will be the geometry column of us.bound. Our function will be *st_centroid()*. We will be using map_dbl variation of map from the purrr package. For more documentation, check out map documentation

To get our longitude values we map the *st_centroid()* function over the geometry column of us.bound and access the longitude value through double bracket notation \[\[\]\] and 1. This allows us to get only the longitude, which is the first value in each centroid.\

```{r}
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])
```

We do the same for latitude with one key difference. We access the second value per each centroid with \[\[2\]\].

```{r}
latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])
```

Now that we have latitude and longitude, we use cbind to put longitude and latitude into the same object.

```{r}
coords <- cbind(longitude, latitude)
```

### DETERMINE THE CUT-OFF DISTANCE

Firstly, we need to determine the upper limit for distance band by using the steps below:

-   Return a matrix with the indices of points belonging to the set of the k nearest neighbours of each other by using [*knearneigh()*](https://r-spatial.github.io/spdep/reference/knearneigh.html) of **spdep**.

-   Convert the knn object returned by *knearneigh()* into a neighbours list of class nb with a list of integer vectors containing neighbour region number ids by using [*knn2nb()*](https://r-spatial.github.io/spdep/reference/knn2nb.html).

-   Return the length of neighbour relationship edges by using [*nbdists()*](https://r-spatial.github.io/spdep/reference/nbdists.html) of **spdep**. The function returns in the units of the coordinates if the coordinates are projected, in km otherwise.

-   Remove the list structure of the returned object by using [**unlist()**](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/unlist).

```{r}
#coords <- coordinates(hunan)
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

The summary report shows that the largest first nearest neighbour distance is 61.79 km, so using this as the upper threshold gives certainty that all units will have at least one neighbour.

### COMPUTING FIXED DISTANCE WEIGHT MATRIX

Now, we will compute the distance weight matrix by using [*dnearneigh()*](https://r-spatial.github.io/spdep/reference/dnearneigh.html) as shown in the code chunk below.

```{r}
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
```

Next, *nb2listw()* is used to convert the nb object into spatial weights object.

```{r}
wm62_lw <- nb2listw(wm_d62, style = 'B')
summary(wm62_lw)
```

The output spatial weights object is called `wm62_lw`.

### COMPUTING ADAPTIVE DISTANCE WEIGHT MATRIX

One of the characteristics of fixed distance weight matrix is that more densely settled areas (usually the urban areas) tend to have more neighbours and the less densely settled areas (usually the rural counties) tend to have lesser neighbours. Having many neighbours smoothes the neighbour relationship across more neighbours.

It is possible to control the numbers of neighbours directly using k-nearest neighbours, either accepting asymmetric neighbours or imposing symmetry as shown in the code chunk below.

```{r}
knn <- knn2nb(knearneigh(coords, k=8))
knn
```

Next, *nb2listw()* is used to convert the nb object into spatial weights object.

```{r}
knn_lw <- nb2listw(knn, style = 'B')
summary(knn_lw)
```

# 7. COMPUTING GI STATISTICS

## 7.1 Gi Statistics Using Fixed Distance

```{r}
fips <- order(hunan$County)
gi.fixed <- localG(hunan$GDPPC, wm62_lw)
gi.fixed
```

The output of localG() is a vector of G or Gstar values, with attributes “gstari” set to TRUE or FALSE, “call” set to the function call, and class “localG”.

The Gi statistics is represented as a Z-score. Greater values represent a greater intensity of clustering and the direction (positive or negative) indicates high or low clusters.

Next, we will join the Gi values to their corresponding hunan sf data frame by using the code chunk below.

```{r}
hunan.gi <- cbind(hunan, as.matrix(gi.fixed)) %>%
  rename(gstat_fixed = as.matrix.gi.fixed.)
```

In fact, the code chunk above performs three tasks. First, it convert the output vector (i.e. *gi.fixed*) into r matrix object by using *as.matrix()*. Next, *cbind()* is used to join hunan\@data and *gi.fixed* matrix to produce a new SpatialPolygonDataFrame called *hunan.gi*. Lastly, the field name of the gi values is renamed to *gstat_fixed* by using *rename()*.

### 7.2 Mapping Gi Values with Fixed Distance Weights

The code chunk below shows the functions used to map the Gi values derived using fixed distance weight matrix.

```{r}
gdppc <- qtm(hunan, "GDPPC")

Gimap <-tm_shape(hunan.gi) +
  tm_fill(col = "gstat_fixed", 
          style = "pretty",
          palette="-RdBu",
          title = "local Gi") +
  tm_borders(alpha = 0.5)

tmap_arrange(gdppc, Gimap, asp=1, ncol=2)
```

> Question: What statistical observation can you draw from the Gi map above?

## **7.3 Gi Statistics Using Adaptive Distance**

The code chunk below are used to compute the Gi values for GDPPC2012 by using an adaptive distance weight matrix (i.e *knb_lw*).

```{r}
fips <- order(hunan$County)
gi.adaptive <- localG(hunan$GDPPC, knn_lw)
hunan.gi <- cbind(hunan, as.matrix(gi.adaptive)) %>%
  rename(gstat_adaptive = as.matrix.gi.adaptive.)
```

## 7.4 Mapping Gi Values with Adaptive Distance Weights

It is time for us to visualise the locations of hot spot and cold spot areas. The choropleth mapping functions of **tmap** package will be used to map the Gi values.

The code chunk below shows the functions used to map the Gi values derived using fixed distance weight matrix.

```{r}
gdppc<- qtm(hunan, "GDPPC")

Gimap <- tm_shape(hunan.gi) + 
  tm_fill(col = "gstat_adaptive", 
          style = "pretty", 
          palette="-RdBu", 
          title = "local Gi") + 
  tm_borders(alpha = 0.5)

tmap_arrange(gdppc, 
             Gimap, 
             asp=1, 
             ncol=2)
```

> Question: What statistical observation can you draw from the Gi map above?
